Closing the gap between statistical and scientific workflows for improved inference in ecology

### 2025-07-10 revision/response_r1.tex

\begin{mybox}
\emph{This opinion manuscript (MS) argues that adapting a modern statistical workflow will help resolve some key issues in forecasting models used in ecology to inform policy about biodiversity loss and climate changes impacts. Specifically, the authors identity two specific “disconnected paradigms” for making such forecasts and argue that a workflow adopted from the Bayesian literature, and that emphasizes data simulation, would help unify these paradigms and make for a better scientific understanding of these important issues. I found the MS well-written, concise and easy to read, and well-organized with clearly laid out challenges and understandable solutions. I agree with the central proposal that the adoption of a “principled” workflow would be beneficial in many areas of ecology, and these types of studies in particular. I have no experience with these specific types of ecological models, but I do have experience with the Bayesian workflow underlying their key recommendation. So, I will focus my review on the recommendation and not the challenge. Below I outline several major concerns/questions, and below that minor comments and corrections.}
\end{mybox}

Response.

\begin{mybox}
\emph{\textbf{Inadequate justification of simulation.}\\
Simulating data and fitting models is key to the step-by-step workflow but I think the MS lacks sufficient detail and motivation. The methods described come from the Bayesian literature, where simulating parameters from priors (prior predictive/pushforward) or posterior (posterior predictive or retrodictive) and then data, are natural. But a quick peek at the cited literature suggests that Bayesian models are not used. So how would one adapt them?}
\end{mybox}

While Bayesian models are more and more frequently (we cite for example ), we clarified that some terms are specific to Bayesian inference but that data simulation still apply for all (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).

@@

Done. % there were two 'then", I removed both...

\begin{mybox}
\emph{Fig 2: Caption should mention this is expanded from Fig. 1C.}
\end{mybox}

\begin{mybox}
\emph{L116, L168: I disagree that “reduce uncertainties” is the goal – instead it is to more accurately quantify it. See L 257 “properly accounting for their uncertainty” for instance.}
\end{mybox}

We have modified ``reduce" to ``quantify" (\href{file:forecastflows_r1\#lintarget:quant1}{line \ref*{lin:quant1}}, \href{file:forecastflows_r1\#lintarget:quant2}{line \ref*{lin:quant2}}).

@@

\begin{mybox}
\emph{L143: FYI the fisheries literature has used the terms pre-data post-model etc. for decades now and have implemented this simulation approach. E.g., https://journal.iwc.int/index.php/jcrm/article/view/239/17. I don’t think it’s necessary to cite this but worth being aware of. You might want to re-cite Hilborn and Mangel (1997) here as they explicitly call for more simulation testing in ecology.}
\end{mybox}

\begin{mybox}
\emph{L201: “..workflow, non…” This sentence partially answers my question about L103 above}
\end{mybox}

\begin{mybox}
\emph{Box A: Please check the logic of the first sentence. More papers because of less time confronting their models? Is that what you meant to say?}
\end{mybox}

\begin{mybox}
\emph{“For examples” should be “For example”?}
\end{mybox}

% Hmm, not sure? "For examples using LPI data" and then we talk about several examples right?

@@

\begin{mybox}
\emph{Box B: Please clarify the last sentence if possible. I’m not sure what “inappropriate way to accommodate their complexity” means here.}
\end{mybox}

We have claried this sentence:
\begin{quote}
``The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model. Calibrating submodels separately allows to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use."
\end{quote}

@@

\begin{mybox}
\emph{L225: I like this paragraph and it makes a point that I hadn’t seen written out so clearly.}
\end{mybox}

Thank you!

\begin{mybox}
\emph{L245: I feel like section 4.2 is slightly out of place. I’d rather see more text devoted to retrodictive checks than this. Consider dropping it and focusing on just a conclusion here}
\end{mybox}

Response. %hmm, he's making a good point, maybe we should at least change the section title?

@@

{\bf Reviewer 2}

\begin{mybox}
\emph{The paper is well written and in a style that is accessible to a wide audience. The topic fits under the theme of “workflow for applied data analysis” by proposing a workflow that if used thoughtfully could lead to less divergence in forecasts and therefore more trust in the scientific process. The proposed workflow is similar to the Bayesian workflow outlined in Gelman et al. (2020) and is translated to two different approaches to ecological forecasting. I think the proposed workflow could be applied or modified to be used with a wide range of research questions within ecological forecasting if careful attention is paid to nuances and idiosyncrasies associated with each unique research question, and I think this necessary attention to question-specific nuances could be highlighted more in the paper. There are places where I feel more specificity (or clarity) with respect to these nuances could improve the manuscript and its potential for impact with respect to building trust in the scientific process. There is also a lack of specificity in how, exactly, this type of workflow could be implemented because there
is no example application to data. I have outlined these areas under “Major comments” and provided minor editorial comments in “Minor Comments.”}
\end{mybox}

% Note: I am an applied statistician with experience collaborating with ecologists and developing
% statistical methodology for large-scale long-term monitoring programs. The bulk of my work
% is more concerned with explanatory than predictive inference and my knowledge of building
% ecological forecasting models for inference on a global scale is limited. I reviewed the paper
% through this lens.

\begin{mybox}
\emph{\textbf{Major comments}\\
1. Clarity on inferential goals of “disconnected paradigms”\\
 When the paper discusses “harmonizing both trend estimation and forecasting” what is meant by this? The title of the paper suggests the inferential goals from both approaches to be predictive in nature (i.e., “forecasting trends”), but it seems plausible that the simpler models are developed for more of an explanatory purpose (e.g., investigating relationships between covariates and mean response) rather than predicting or forecasting future trends, which would make the questions being asked by each modeling approach fundamentally different (see, Shmueli 2010). The paper could be improved if more clarity was provided on the inferential goals of both types of modeling paradigms it discusses, and perhaps provide some discussion why they need to be (or if they should be) harmonized if they are in pursuit of different types of inference.}
\end{mybox}

Thank you for this suggestion. We added a paragraph in the introduction to discuss the inferential goals of these two approachs (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}). We also changed the title to reflect the broader inferential goals.

@@

\begin{mybox}
\emph{2. Can this workflow be applied to non-Bayesian models\\
The emphasis on simulation-based investigation of model components is highlighted as a key feature of
the workflow, so it seems a generative (or at least partially generative) model is assumed. Does this model need to be Bayesian? Explicit connections to posterior predictive checks and more discussion for how both prior- and posterior-predictive assessment could be used in a non-Bayesian context would strengthen this work. (Also note that prior predictive checks should be made explicit too – see minor comments).}
\end{mybox}

We clarified that some terms are specific to Bayesian inference but that this workflow can be applied more broadly (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).

\begin{mybox}
\emph{3. Discussion of Calibration:\\
Lines 98-104 seem to be referring to a check on estimating “the model” using synthetic data that are generated from the model (either through optimization or McMC), but there is no mention of various computational approaches that may be taken to fit “the model.” How does the chosen computational strategy fit into the workflow if at all?}
\end{mybox}

While we agree that this is an interesting question, we believe it falls outside the scope of this manuscript.

\begin{mybox}
\emph{Further, is the proposed workflow to conduct a full-blown simulation calibration study (i.e., simulate-fit-check for capture) many times and across all possible parameter values? Assuming yes, I’d be interested in how something like that would scale for the types of models that are the focus of the paper? Could the authors provide more guidance on how to conduct such a study in this context?}
\end{mybox}

We have clarified the data simulaton step. We believe a first step would be to simulate from one parameter set (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}), and then to repeat the simulations with several parameter sets (\href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}).

@@

\begin{mybox}
\emph{4. Gaps in the workflow\\
A big argument in the paper is that the proposed workflow could lessen discrepancies in results between studies of trends (lines 123-128). However, in the workflow there is no mention of transparent communication of study design, which if ignored in analysis, could result in discrepancies in results for studies with similar research questions. Differences in sampling designs must also be considered prior tointegrating different data sources (if integration is even possible). The importance of study design/sampling design and its ties to appropriate modeling is not discussed. Is itimplicitly assumed that there is alignment between design, question, and model in this workflow? Some mention of the importance of study design would strengthen the work.}
\end{mybox}

Response.

\begin{mybox}
\emph{Additionally, the choice of prior and posterior predictive checks (PPC) is non-trivial and depends on the context of the study. There is discussion of these types of simulation-based investigations, but no mention of the importance of developing checks that assess key features of the observed data that are tied to the research question. Acknowledging the complexity of this step, and including some ideas for how to do this in the context of the types of problems discussed in the paper could strengthen the paper.}
\end{mybox}

Thank you for this suggestion. We now say that the checks should be developped specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).

@@

\begin{mybox}
\emph{Figure 1 caption, line 34: change “where a model is build” to “where a model is built”}
\end{mybox}

Done.

\begin{mybox}
\emph{Lines 75-81, the paper discusses a form of prior predictive checking without explicitly mentioning the connection. The connection should be mentioned and perhaps}
\end{mybox}

We now mention that this step can allow to check our prior assumptions (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}).

@@

\begin{mybox}
\emph{Line 86: change “also sometimes called a posterior predictive check” to “in a Bayesian workflow this is called a posterior predictive check.”}
\end{mybox}

...

\begin{mybox}
\emph{Line 87: could be more specific about “model output” as typically a PPC investigates how the entire posterior predictive distribution for the response or some summary statistic that is a function of the response and parameters compares to the observed data (or observed summary stat computed from observed data).}
\end{mybox}

We now mention that the summary statistics should be developped specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}). % I already said this above...

@@

\begin{mybox}
\emph{Line 117: should “hypothesis” be “hypotheses”?}
\end{mybox}

Corrected.

\begin{mybox}
\emph{Lines 198-201: if the goal is accurate and relatively-precise forecasts, do all of the nuisance parameters need to be identifiable for a model to be useful? That is, does identifiability of all parameters matter for process-based models with a goal of forecasting?}
\end{mybox}

We believe that the strength of process-based models (as compared to machine learning for example, \href{file:forecastflows_r1\#lintarget:MLvsPBM}{line \ref*{lin:MLvsPBM}}) is their interpretability. Their parameters should represent biological processes, and their values are of interest. The inferential goal of process-based models is both predictive and explanatory (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}), and poor identifiability could reduce this.

### 2025-07-10 revision/forecastflows_r1.tex
continue working on reviewer response

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
\date{\today}
\author{Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
\maketitle
\noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}\\

\noindent \emph{For:} Workflow for Applied Data Analysis theme issue for \emph{Phil Trans A} as an \emph{Opinion}
% They mention a tex template here: https://royalsocietypublishing.org/rsta/for-authors#question4
% Word length: I should check what they emailed but they say no more than 13 printed pages with 650 per page (whoa, we should be well below that! I am thinking 3-5K would be good)
% Already 5K!

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models. Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons. First, the research community is split between two disconnected paradigms: using data to fit simple, trend-like models with few parameters, or developing forecasting models that include complex, mechanistic submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.

@@

Anthropogenic drivers are reshaping natural systems \citep{Diaz2019}. Impacts are projected to increase in coming decades, as climate change accelerates biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}.
Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics. % These requirements in turn have created a growing need for more global ecological data, and the types of models that can robustly handle such data.

@@

We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects; the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components. These new parts are often disconnected from the original research aim, its data stream, and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).

Workflows that fully integrate all the steps required to build a model from an ecological question, with evaluation of limitations and potential problems before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge \citep{McElreath2018, betanworkflow, Gelman2020, Wolkovich2024}. Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

\begin{figure}[h]
	\centering
	\includegraphics{../figures/modeldataspaces_rotation}
	\caption{Different trajectories through model-data space during model development. Ideally, this begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is build---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedbacks to refine our model.}
	\label{fig:modeldata}
\end{figure}

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}a). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (figure \ref{fig:modeldata}b). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

We argue a workflow that moves along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error.  
\llabel{causalinf}This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}. From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}). The general idea is to start with a relatively simple model that we could refine later (see example worklow we provide). At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable values (which is straightforward if the parameters are interpretable) and from fake predictor data. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is reasonable or not. This is also a step for making sure the model is working as expected (see example worklow). \llabel{datasimforall}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range (or, in a Bayesian framework, from the prior distribution).

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations. 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations. 
It's only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

@@

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es) (if we can identify them) and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties and yield new scientific insights. This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.

@@

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. 
If a model becomes too complex, it may become a black box,
and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

@@

With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important. Machine learning will likely surpass process-based models for forecasting accuracy if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability. Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.

>>>>

% \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
\title{Closing the gap between statistical and scientific workflows for improved inference in ecology} 
\date{\today}
\author{Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
\maketitle
\noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}\\

\noindent \emph{For:} Workflow for Applied Data Analysis theme issue for \emph{Phil Trans A} as an \emph{Opinion}

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models. Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons. First, the research community is split between two disconnected paradigms: using data to fit simple, trend-like models with few parameters, or developing forecasting models that include complex, mechanistic submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform ecological modeling.

@@

Anthropogenic drivers are reshaping natural systems \citep{Diaz2019}. Impacts are projected to increase in coming decades, as climate change accelerates biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}.
Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date, alongside the skill to forecast future dynamics. % These requirements in turn have created a growing need for more global ecological data, and the types of models that can robustly handle such data.

@@

\llabel{inferentialgoal}The inferential goals of these two modeling paradigms---explanation vs.\ prediction---seem disconnected at first glance, but remains in reality much more blurred. Trend estimation, often with simpler models, tend to focus on an explanatory purpose, but often lacks a clear link with theories that could provide causal interpretation. For forecasting, the use of more complex process-based models for a predictive purpose is often justified by the assumption that ``we must understand to predict", contributing to the confusion between explanation and prediction \citep{Shmueli2010}.

We argue that current debates, confusions and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects; the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components. These new parts are often disconnected from the original research aim, its data stream, and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).

Workflows that fully integrate all the steps required to build a model from an ecological question, with evaluation of limitations and potential problems before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge \citep{McElreath2018, betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024}. Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

\begin{figure}[h]
	\centering
	\includegraphics{../figures/modeldataspaces_rotation}
	\caption{Different trajectories through model-data space during model development. Ideally, this begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedbacks to refine our model.}
	\label{fig:modeldata}
\end{figure}

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}a). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (figure \ref{fig:modeldata}b). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

We argue a workflow that moves along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error.  
\llabel{causalinf}This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}. From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}). The general idea is to start with a relatively simple model that we could refine later (see example workflow we provide). At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).

With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}). `Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values (which is straightforward if the parameters are interpretable) and from fake predictor data. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not. This is also a step for making sure the model is working as expected (see example worklow). \llabel{repeatsim}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range. \llabel{priorPC}In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors, \citealp{Gelman2020}).

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}). This way, we obtain parameter estimates constrained by observations. 
These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}). This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations. \llabel{howtoPPC}There is no general way to do this: the goal is to tailor checks that assess key features of the data in regards to our research question \citep{Gelman2020}.
It's only once all steps have been completed that we can interpret parameter values with respect to our research question.
The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.

@@

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops---\llabel{broadcontext}and should be applied not only in a Bayesian framework but across broader modeling approaches.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es) (if we can identify them) and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}quantify uncertainties and yield new scientific insights. This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data. Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.

@@

Increasing model complexity can be beneficial, especially when it helps better \llabel{quant2}quantify the uncertainty; however, this is not always the outcome. In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. 
If a model becomes too complex, it may become a black box,
and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

@@

With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important. \llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability. Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.
